Post-Disturbance Tree Community Trajectories in a Neotropical Forest

### 2019-02-09 WholePlotTrajectories.Rmd
intro

The large areas covered with tropical forests worldwide hold crucial environmental, economic and social values. 
They provide wood and multiple non-timber forest products, shelter a diversified fauna, and ensure cultural and human well-being.
They regulate as well the local and regional climates, and the carbon, water and nutrient cycles.
However, the growing demand in forests products together with current global changes increase the pressure on remaining natural forests [@Morales-Hidalgo2015].
This growing pressure affects the natural disturbance regime that shape and maintain the structure, composition and functioning of tree communities [@Schnitzer2001;@Anderson-Teixeira2013;@Sist2015].
It is then urgent to understand the processes underlying community response to disturbance in order to anticipate the fate of tropical forests in the current context.
In tropical forests, disturbance maintains a forest environment and is followed by secondary succession comprising a suit of changes in the structure and functioning of communities.
Disturbance change both community abiotic and biotic environments, through modifications in the fluxes of light, heat and water [@Goulamoussene2017]. 
For now, post-disturbance trajectories has been largely studied through forest structural parameters such as aboveground biomass, tree height or stem density [@Piponiot2016;@Rutishauser2016]. 
Models based on observed trajectories then assessed community time to recovery and highlighted the determinants of post-disturbance trajectories, as forest structure, forest composition, environmental parameters and disturbance characteristics [@Herault2018]. 
Regarding community diversity and composition, however, post-disturbance trajectories have not been as thoroughly understood [@Guitet2018;@Molino2001].
Biodiversity trajectories result from the differences in species response to disturbance which entail a variety of processes like colonisation, facilitation, coexistence [@Lindenmayer2012;@Garcia_florez2017].
The rate, endpoint and duration of post-disturbance trajectories depend on the intensity and frequency of disturbance: a current objective of tropical ecology is to highlight this linkage betwenn biodiversity and disturbance, and anticipate community response [@Chazdon2003a;@Willig2018].

An early conceptual basis of the linkage between biodiversity and disturbance is the Intermediate Disturbance Hypothesis (IDH).
The IDH assumes a relationship between community diversity and the intensity and frequency of disturbance events [@Connell1978], and states that community diversity peaks at intermediate level of disturbance.
The fluctuations of environmental conditions after disturbance are assumed to prevent competitive exclusion as they foster both competitively superior species and fast colonizers [@Shea2004;@Pulsford2016].
In highly diverse tropical forests, however, observations of the IDH are often blurry compared to theoretical expectations [@Hugues2007;@Sheil2003;@Norden2017] the validation of the theory remains controversial [@Hubbell2001;@Fox2013;@Sheil2013].
As previously highlighted, there is a variety of ecological processes emerging after disturbance, and the pattern of the IDH (the diversity peak under intermediate disturbance) can result from various processes.
Disentangle these processes in the context of tropical forests would assess the conditions for the IDH application and clarify the disturbance impact on communit diversity [@Sheil2003;@Shea2004].

Empirical evidence of the IDH generally rely on the analysis of a single diversity metric, often the taxonomic richness [@Molino2001], which is limited as neither community changes in species abundance nor in functional characteristics are accounted for [@Mayfield2010;@Villeger2011]. 
The post-disturbance processes involved, though, can only be disentangled in combining different approaches [@Sheil2003;@Shea2004].
More ecological-meaningful analysis would couple taxonomic richness, evenness, and composition, to highlight the changes in species composition and abundance distribution, and hence the changes in fostered species [@Chaudhary2016]. 
Analysis would as well include a functional approach, that accounts for species biological attributes, would include as the trait-based approach indeed directly links community diversity, composition and redundancy to ecosystem functioning and environmental constraints [@Violle2007b;@Baraloto2012a].

In that respect, a vast literature allowed recognizing major traits that represent species ecological strategy and determine how they respond to changing conditions [@Diaz2005]. 
Specifically in tropical forests, the functional approach revealed the post-disturbance deterministic processes entailing a shift from a dominance of “conservative” slow-growing species dealing with scarce resources, to a dominance of “acquisitive” fast-growing species with rapid and efficient use of abundant resources [@Rees2001;@Reich2014;@Herault2011]. 
This shift is translated into the trajectories of average community value of key functional traits related to resource acquisition (leaf and stem traits) and life-history strategy (seed mass, maximum size)  [@Wright2004;@TerSteege2006;@Westoby2006a;@Chave2009b]. 
The functional approach also encompasses the analysis of functional redundancy, that quantifies the amount of shared trait values among species [@Carmona2016]. The high functional redundancy of hyper-diverse tropical forests [@Bellwood2006] mitigates the impacts of species removal on ecosystem functioning and determines the resilience of communities after disturbance [@Elmqvist2003; @Diaz2005].

In this study, we monitored over 30 years the response of 75 ha of Neotropical forest plots set up on a gradient of disturbance intensity, from 10 to 60% of above-ground biomass (AGB) loss. We made use of a large functional traits database encompassing major leaf, stem and life-history traits in order to draw the taxonomic and functional trajectories in terms of richness, evenness, composition and redundancy. Specifically, we elucidateed (i) the community taxonomic and functional recovery and (ii) the ecological processes underlying community post-disturbance trajectories and discussed the validity and the extent of the IDH, and (iii) examined community recovery time.

>>>>

The large areas covered with tropical forests worldwide hold crucial environmental, economic and social values. 
They provide wood and multiple non-timber forest products, shelter a diversified fauna, and ensure cultural and human well-being.
They regulate as well the local and regional climates, and the carbon, water and nutrient cycles.
However, the growing demand in forests products together with current global changes increase the pressure on remaining natural forests [@Morales-Hidalgo2015].
These threats affect the natural disturbance regime that shape and maintain the structure, composition and functioning of tree communities [@Schnitzer2001;@Anderson-Teixeira2013;@Sist2015].
It is now urgent to understand the processes underlying community response to disturbance, and thus to be able to anticipate the fate of tropical forests in the current context.
In tropical forests, disturbance maintains a forest environment and initiates a suit of changes in the structure and functioning of communities consituting a secondary succession.
Disturbance change both community abiotic and biotic environments, through modifications in the fluxes of light, heat and water [@Goulamoussene2017]. 
For now, post-disturbance trajectories has been largely studied through forest structural parameters such as aboveground biomass, tree height or stem density [@Piponiot2016;@Rutishauser2016]. 
Models based on observed trajectories then assessed community time to recovery and identified some of the determinants of post-disturbance trajectories, like forest structure, forest composition, environmental parameters and disturbance characteristics [@Herault2018]. 
Regarding community diversity and composition, however, post-disturbance trajectories have not been as thoroughly understood [@Guitet2018;@Molino2001].
The variety of species response to disturbance entail different processes like colonisation, facilitation, coexistence, etc, which succession shapes biodiversity post-disturbance trajectories [@Lindenmayer2012;@Garcia_florez2017].
The rate, endpoint and duration of the trajectories depend on the intensity and frequency of disturbance: a current objective of tropical ecology is to highlight the link betwenn biodiversity and disturbance and anticipate community trajectories [@Chazdon2003a;@Willig2018].

An early conceptual basis of the linkage between biodiversity and disturbance is the Intermediate Disturbance Hypothesis (IDH).
The IDH assumes a relationship between community diversity and the intensity and frequency of disturbance events [@Connell1978].
The theory and states that community diversity peaks at intermediate level of disturbance.
The fluctuations of environmental conditions after disturbance are assumed to prevent competitive exclusion in fostering both competitively superior species and fast colonizers [@Shea2004;@Pulsford2016].
In highly diverse tropical forests, however, observations of the IDH are often blurry compared to theoretical expectations [@Hugues2007;@Sheil2003;@Norden2017], and the validation of the theory remains controversial [@Hubbell2001;@Fox2013;@Sheil2013].
As previously highlighted, there is a variety of ecological processes emerging after disturbance, and the pattern of the IDH (the diversity peak under intermediate disturbance) can result from several such processes.
Disentangle the processes involved and their succession in time would allow discussing the conditions for the application of the IDH, its scope in time, and its implication for the fate of forests [@Sheil2003;@Shea2004].

Examine post-disturbance trajectories requires a set of diversity metrics, to grasp all aspects of community changes [@Sheil2003;@Shea2004;@Mayfield2010].
Specifically, diversity metrics should assess taxonomic richness, evenness and composition, in order to highlight the changes the community abundance distribution and in the species fostered by the disturbance. 
Analysis would as well include a functional approach accounting for species biological attributes and ecological strategy.
The trait-based approach directly links community diversity, composition and redundancy to ecosystem functioning and environmental constraints [@Violle2007b;@Baraloto2012a]. 
In that respect, a vast literature allowed recognizing major traits that represent species ecological strategy and determine how they respond to changing conditions [@Diaz2005]. 
Specifically in tropical forests, the functional approach revealed the post-disturbance deterministic processes entailing a shift from a dominance of “conservative” slow-growing species dealing with scarce resources, to a dominance of “acquisitive” fast-growing species with rapid and efficient use of abundant resources [@Rees2001;@Reich2014;@Herault2011]. 
This shift is translated into the trajectories of average community value of key functional traits related to resource acquisition (leaf and stem traits) and life-history strategy (seed mass, maximum size)  [@Wright2004;@TerSteege2006;@Westoby2006a;@Chave2009b]. 
The functional approach also encompasses the analysis of functional redundancy, that quantifies the amount of shared trait values among species [@Carmona2016]. The high functional redundancy of hyper-diverse tropical forests [@Bellwood2006] mitigates the impacts of species removal on ecosystem functioning and determines the resilience of communities after disturbance [@Elmqvist2003; @Diaz2005].

In this study, we monitored over 30 years the response of 75 ha of Neotropical forest plots set up on a gradient of disturbance intensity, from 10 to 60% of above-ground biomass (AGB) loss. We made use of a large functional traits database encompassing major leaf, stem and life-history traits in order to draw the taxonomic and functional trajectories in terms of richness, evenness, composition and redundancy. Specifically, we (i) elucidated community taxonomic and functional resilience, (ii) discussed the scope of the IDH and the underlying ecological processes, and (iii) examined community recovery time.

### 2019-02-09 WholePlotTrajectories.tex
intro

The large areas covered with tropical forests worldwide hold crucial
environmental, economic and social values. They provide wood and
multiple non-timber forest products, shelter a diversified fauna, and
ensure cultural and human well-being. They regulate as well the local
and regional climates, and the carbon, water and nutrient cycles.
However, the growing demand in forests products together with current
global changes increase the pressure on remaining natural forests
\citep{Morales-Hidalgo2015}. This growing pressure affects the natural
disturbance regime that shape and maintain the structure, composition
and functioning of tree communities
\citep{Schnitzer2001, Anderson-Teixeira2013, Sist2015}. It is then
urgent to understand the processes underlying community response to
disturbance in order to anticipate the fate of tropical forests in the
current context. In tropical forests, disturbance maintains a forest
environment and is followed by secondary succession comprising a suit of
changes in the structure and functioning of communities. Disturbance
change both community abiotic and biotic environments, through
modifications in the fluxes of light, heat and water
\citep{Goulamoussene2017}. For now, post-disturbance trajectories has
been largely studied through forest structural parameters such as
aboveground biomass, tree height or stem density
\citep{Piponiot2016, Rutishauser2016}. Models based on observed
trajectories then assessed community time to recovery and highlighted
the determinants of post-disturbance trajectories, as forest structure,
forest composition, environmental parameters and disturbance
characteristics \citep{Herault2018}. Regarding community diversity and
composition, however, post-disturbance trajectories have not been as
thoroughly understood \citep{Guitet2018, Molino2001}. Biodiversity
trajectories result from the differences in species response to
disturbance which entail a variety of processes like colonisation,
facilitation, coexistence \citep{Lindenmayer2012, Garcia_florez2017}.
The rate, endpoint and duration of post-disturbance trajectories depend
on the intensity and frequency of disturbance: a current objective of
tropical ecology is to highlight this linkage betwenn biodiversity and
disturbance, and anticipate community response
\citep{Chazdon2003a, Willig2018}.

An early conceptual basis of the linkage between biodiversity and
disturbance is the Intermediate Disturbance Hypothesis (IDH). The IDH
assumes a relationship between community diversity and the intensity and
frequency of disturbance events \citep{Connell1978}, and states that
community diversity peaks at intermediate level of disturbance. The
fluctuations of environmental conditions after disturbance are assumed
to prevent competitive exclusion as they foster both competitively
superior species and fast colonizers \citep{Shea2004, Pulsford2016}. In
highly diverse tropical forests, however, observations of the IDH are
often blurry compared to theoretical expectations
\citep{Hugues2007, Sheil2003, Norden2017} the validation of the theory
remains controversial \citep{Hubbell2001, Fox2013, Sheil2013}. As
previously highlighted, there is a variety of ecological processes
emerging after disturbance, and the pattern of the IDH (the diversity
peak under intermediate disturbance) can result from various processes.
Disentangle these processes in the context of tropical forests would
assess the conditions for the IDH application and clarify the
disturbance impact on communit diversity \citep{Sheil2003, Shea2004}.

Empirical evidence of the IDH generally rely on the analysis of a single
diversity metric, often the taxonomic richness \citep{Molino2001}, which
is limited as neither community changes in species abundance nor in
functional characteristics are accounted for
\citep{Mayfield2010, Villeger2011}. The post-disturbance processes
involved, though, can only be disentangled in combining different
approaches \citep{Sheil2003, Shea2004}. More ecological-meaningful
analysis would couple taxonomic richness, evenness, and composition, to
highlight the changes in species composition and abundance distribution,
and hence the changes in fostered species \citep{Chaudhary2016}.
Analysis would as well include a functional approach, that accounts for
species biological attributes, would include as the trait-based approach
indeed directly links community diversity, composition and redundancy to
ecosystem functioning and environmental constraints
\citep{Violle2007b, Baraloto2012a}.

In that respect, a vast literature allowed recognizing major traits that
represent species ecological strategy and determine how they respond to
changing conditions \citep{Diaz2005}. Specifically in tropical forests,
the functional approach revealed the post-disturbance deterministic
processes entailing a shift from a dominance of ``conservative''
slow-growing species dealing with scarce resources, to a dominance of
``acquisitive'' fast-growing species with rapid and efficient use of
abundant resources \citep{Rees2001, Reich2014, Herault2011}. This shift
is translated into the trajectories of average community value of key
functional traits related to resource acquisition (leaf and stem traits)
and life-history strategy (seed mass, maximum size)
\citep{Wright2004, TerSteege2006, Westoby2006a, Chave2009b}. The
functional approach also encompasses the analysis of functional
redundancy, that quantifies the amount of shared trait values among
species \citep{Carmona2016}. The high functional redundancy of
hyper-diverse tropical forests \citep{Bellwood2006} mitigates the
impacts of species removal on ecosystem functioning and determines the
resilience of communities after disturbance
\citep{Elmqvist2003, Diaz2005}.

In this study, we monitored over 30 years the response of 75 ha of
Neotropical forest plots set up on a gradient of disturbance intensity,
from 10 to 60\% of above-ground biomass (AGB) loss. We made use of a
large functional traits database encompassing major leaf, stem and
life-history traits in order to draw the taxonomic and functional
trajectories in terms of richness, evenness, composition and redundancy.
Specifically, we elucidateed (i) the community taxonomic and functional
recovery and (ii) the ecological processes underlying community
post-disturbance trajectories and discussed the validity and the extent
of the IDH, and (iii) examined community recovery time.

>>>>

The large areas covered with tropical forests worldwide hold crucial
environmental, economic and social values. They provide wood and
multiple non-timber forest products, shelter a diversified fauna, and
ensure cultural and human well-being. They regulate as well the local
and regional climates, and the carbon, water and nutrient cycles.
However, the growing demand in forests products together with current
global changes increase the pressure on remaining natural forests
\citep{Morales-Hidalgo2015}. These threats affect the natural
disturbance regime that shape and maintain the structure, composition
and functioning of tree communities
\citep{Schnitzer2001, Anderson-Teixeira2013, Sist2015}. It is now urgent
to understand the processes underlying community response to
disturbance, and thus to be able to anticipate the fate of tropical
forests in the current context. In tropical forests, disturbance
maintains a forest environment and initiates a suit of changes in the
structure and functioning of communities consituting a secondary
succession. Disturbance change both community abiotic and biotic
environments, through modifications in the fluxes of light, heat and
water \citep{Goulamoussene2017}. For now, post-disturbance trajectories
has been largely studied through forest structural parameters such as
aboveground biomass, tree height or stem density
\citep{Piponiot2016, Rutishauser2016}. Models based on observed
trajectories then assessed community time to recovery and identified
some of the determinants of post-disturbance trajectories, like forest
structure, forest composition, environmental parameters and disturbance
characteristics \citep{Herault2018}. Regarding community diversity and
composition, however, post-disturbance trajectories have not been as
thoroughly understood \citep{Guitet2018, Molino2001}. The variety of
species response to disturbance entail different processes like
colonisation, facilitation, coexistence, etc, which succession shapes
biodiversity post-disturbance trajectories
\citep{Lindenmayer2012, Garcia_florez2017}. The rate, endpoint and
duration of the trajectories depend on the intensity and frequency of
disturbance: a current objective of tropical ecology is to highlight the
link betwenn biodiversity and disturbance and anticipate community
trajectories \citep{Chazdon2003a, Willig2018}.

An early conceptual basis of the linkage between biodiversity and
disturbance is the Intermediate Disturbance Hypothesis (IDH). The IDH
assumes a relationship between community diversity and the intensity and
frequency of disturbance events \citep{Connell1978}. The theory and
states that community diversity peaks at intermediate level of
disturbance. The fluctuations of environmental conditions after
disturbance are assumed to prevent competitive exclusion in fostering
both competitively superior species and fast colonizers
\citep{Shea2004, Pulsford2016}. In highly diverse tropical forests,
however, observations of the IDH are often blurry compared to
theoretical expectations \citep{Hugues2007, Sheil2003, Norden2017}, and
the validation of the theory remains controversial
\citep{Hubbell2001, Fox2013, Sheil2013}. As previously highlighted,
there is a variety of ecological processes emerging after disturbance,
and the pattern of the IDH (the diversity peak under intermediate
disturbance) can result from several such processes. Disentangle the
processes involved and their succession in time would allow discussing
the conditions for the application of the IDH, its scope in time, and
its implication for the fate of forests \citep{Sheil2003, Shea2004}.

Examine post-disturbance trajectories requires a set of diversity
metrics, to grasp all aspects of community changes
\citep{Sheil2003, Shea2004, Mayfield2010}. Specifically, diversity
metrics should assess taxonomic richness, evenness and composition, in
order to highlight the changes the community abundance distribution and
in the species fostered by the disturbance. Analysis would as well
include a functional approach accounting for species biological
attributes and ecological strategy. The trait-based approach directly
links community diversity, composition and redundancy to ecosystem
functioning and environmental constraints
\citep{Violle2007b, Baraloto2012a}. In that respect, a vast literature
allowed recognizing major traits that represent species ecological
strategy and determine how they respond to changing conditions
\citep{Diaz2005}. Specifically in tropical forests, the functional
approach revealed the post-disturbance deterministic processes entailing
a shift from a dominance of ``conservative'' slow-growing species
dealing with scarce resources, to a dominance of ``acquisitive''
fast-growing species with rapid and efficient use of abundant resources
\citep{Rees2001, Reich2014, Herault2011}. This shift is translated into
the trajectories of average community value of key functional traits
related to resource acquisition (leaf and stem traits) and life-history
strategy (seed mass, maximum size)
\citep{Wright2004, TerSteege2006, Westoby2006a, Chave2009b}. The
functional approach also encompasses the analysis of functional
redundancy, that quantifies the amount of shared trait values among
species \citep{Carmona2016}. The high functional redundancy of
hyper-diverse tropical forests \citep{Bellwood2006} mitigates the
impacts of species removal on ecosystem functioning and determines the
resilience of communities after disturbance
\citep{Elmqvist2003, Diaz2005}.

In this study, we monitored over 30 years the response of 75 ha of
Neotropical forest plots set up on a gradient of disturbance intensity,
from 10 to 60\% of above-ground biomass (AGB) loss. We made use of a
large functional traits database encompassing major leaf, stem and
life-history traits in order to draw the taxonomic and functional
trajectories in terms of richness, evenness, composition and redundancy.
Specifically, we (i) elucidated community taxonomic and functional
resilience, (ii) discussed the scope of the IDH and the underlying
ecological processes, and (iii) examined community recovery time.
